Multi-level modeling with Openflexo/\FML

### 2021-09-24 responseLetter.tex

\begin{response}{There are some minor typos throughout the manuscript (eg also regarding the reference Jeusfeld2019; choose -> chose; axis -> axes; aN FML virtual model...; etc.)} 
TODO.
choose $\rightarrow$ chose OK ; axis $\rightarrow$  axes OK ; je ne comprends pas pour Jeusfeld2019
(deeptelos2019) ni pour aN FML (je dirais ``a FML'', mais je puis me tromper)
\end{response}

\pagebreak

\Reviewer{\#2}

\begin{response}{The basic idea of model federation is interesting, its primary focus is not on creating multi-level models. However, as the approach is quite flexible, multi-level behavior can be emulated by model federation. This fact is possibly the most serious advantage and disadvantage of the paper at the same time. Advantage, since it brings us a completely new, different way of thinking, and disadvantage since it does not really help in categorizing/unifying multi-level methods and – for me – it also feels a little bit off-topic in this special issue. At this point, it should be mentioned that the authors managed to cover/solve all requirements of the challenge, thus the paper, as a solution to the challenge fits well.} We introduced at the end of the introduction a description of our approach. In a word, we use FML (linguistic level) as a tool to build models and meta-models that allows multi-level modelling. The evolution of models (multi-levels) is encoded as a hierarchy of models and meta-models. We do multi-level modelling on demands, and propose an organization guided by models evolution. 
\todo{Answering this is going to be tricky!}
\end{response}

### 2021-09-24 9-relatedworks.tex
Extending related work.

%Salvador commence
A plethora of multi-level modeling approaches and tools with different foundations have appeared in recent years\footnote{\url{https://homepages.ecs.vuw.ac.nz/Groups/MultiLevelModeling/MultiTools}}. Comparing them lies out of the scope of this paper. In this sense, we limit the following discussion to the presentation and comparison of previous solutions to the MULTI Process Challenge.

A first description of process modeling as a multi-level modeling problem was proposed by \parencite{lara2018refactoring} in the context of a catalogue of refactoring for multi-level models (in a simpler form with fewer constraints and requirements w.r.t. the challenge version). A solution is provided with MetaDepth~\parencite{metadepth}, which supports modeling with any number of levels, dual ontological/linguistic typing and deep characterization through potency.

\cite{multiecore2019} use MultiEcore \citep{multecore2016} to provide a solution to the challenge. MultEcore uses an extension of the two-level cascading technique and potency. They do not need the synthetic typing relations we introduce. However they need to do frequent model transformations to transform instance models into instantiable models. With respect to the proposed solution, we use fewer models as we do not include a generic software engineering process model nor a generic assurance process model. %Finally, they use multi-level couple model transformations for the definition and enforcement of constraints.

@@

>>>>

%Salvador commence

\textcolor{red}{Our solution to the Multi-Level process challenge relies on the use of the type/object pattern~\citep{typeObject} for achieving ontological instantiation and the core capacity of Openflexo/FML to reference models disregarding abstraction levels. Concretely, we use a metamodel to group the required concepts for describing universal processes and an extending metamodel that adapts it to the domain of software engineering processes. Process instances may reference elements from both metamodels.}

A plethora of multi-level modeling approaches and tools with different foundations have appeared in recent years\footnote{\url{https://homepages.ecs.vuw.ac.nz/Groups/MultiLevelModeling/MultiTools}}. Comparing them lies out of the scope of this paper. In this sense, we limit the following discussion to the presentation and comparison of previous solutions to the MULTI Process Challenge.

A first description of process modeling as a multi-level modeling problem was proposed by \parencite{lara2018refactoring} in the context of a catalogue of refactoring for multi-level models (in a simpler form with fewer constraints and requirements w.r.t. the challenge version). A solution is provided with MetaDepth~\parencite{metadepth}, which supports modeling with any number of levels, dual ontological/linguistic typing and deep characterization through potency~\citep{potency}. \textcolor{red}{For their solution to the multi-level process problem they use three levels. The first level describes generic processes, the second level, software engineering processes and the third level, software engineering enacted processes. The use \emph{linguistic extensions}\citep{metadepth}, a mechanism to  linguistically extend ontological instance models, in order to introduce artefact types and tasks duration in the second level.}

\cite{multiecore2019} use MultEcore \citep{multecore2016} to provide a solution to the challenge. \textcolor{red}{MultEcore uses, (un)pluggable linguistic levels (e.g, there is not a \emph{fixed} core metamodel defining the concept of clabjet), an extension of the two-level cascading technique \citep{atkinson2005concepts} (this is achieved with model transformations to transform instance models into instantiable models) and potency. With respect to the challenge, their solution uses 4 levels. The first level represents generic processes and constitutes the root level for two ontological hierarchies, one for the software engineering domain and the other for the insurance domain. Each hierarchy contains three other ontological levels. The second level refines generic process to software engineering and insurance processes. The third level refines the aforementioned models to adapt them to the ACME and Xsure processes. Finally, process instances lie in level four. Additionally, the authors use an orthogonal linguistic hierarchy to fulfill requirement \textbf{P19} about alternative names. With respect to our solution, the authors use more models, as we require less refinement steps. However accidental complexity is introduced by our approach as we need additional constructs and constraints for the typing relations which are otherwise built-in in MultEcore.} \noteSM{How do we deal with P19?}

@@

\noteSM{Add some infor about how constraints are managed in each of the approaches.}
\noteSM{e.g., built-in constraints for typing relations}
